---
title: Model 01
---

- Ignore partie mergers
- Ignore errors

# Preparation

Load libraries.

```{r packages}
# Data
library(nanoparquet)
library(dplyr)
library(tidyr)
library(stringr)
# Bayes
library(cmdstanr)
library(posterior)
library(bayesplot)
# Ploting
library(ggplot2)
library(ggnewscale)
```

Check Stan installation.

```{r stan-check}
options("mc.cores" = future::availableCores() - 1)
cmdstanr::check_cmdstan_toolchain()
```

ggplot.

```{r ggplot-utils}
scale_color_parties <- function(parties) {
  scale_color_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$party_stan),
    labels = setNames(parties$party_shortcut, parties$party_stan)
  )
}

scale_color_houses <- function(houses) {
  list(
    scale_color_manual(
      name = "Polling House",
      breaks = as.character(house_map$h),
      values = setNames(house_map$color, as.character(house_map$h)),
      labels = setNames(
        as.character(house_map$agency),
        as.character(house_map$h)
      )
    ),
    theme(
      legend.position = "bottom"
    )
  )
}

scale_fill_parties <- function(parties) {
  scale_fill_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$party_stan),
    labels = setNames(parties$party_shortcut, parties$party_stan)
  )
}

scale_x_parties <- function(parties) {
  list(
    scale_x_continuous(
      breaks = parties$party_stan,
      labels = parties$party_shortcut
    ),
    theme(
      panel.grid.major.x = element_blank()
    )
  )
}

scale_x_polls <- function(cal) {
  list(
    scale_x_continuous(
      name = NULL,
      breaks = seq(from = 0, to = max(cal$t), by = 50),
      minor_breaks = seq(from = 0, to = max(cal$t), by = 25),
      limits = c(0, max(cal$t)),
      expand = expansion(add = 5)
    ),
    theme(
      panel.grid.minor.x = element_line(linetype = "dashed")
    )
  )
}

scale_x_houses <- function(houses) {
  list(
    scale_x_continuous(
      breaks = houses$h,
      labels = houses$agency
    ),
    theme(
      panel.grid.major.x = element_blank()
    )
  )
}

scale_y_voteshare <- function(ylim = 1) {
  list(
    scale_y_continuous(
      name = "voteshare [%]",
      breaks = seq(0, 1, by = 0.1),
      minor_breaks = seq(0, 1, by = 0.05),
      labels = scales::label_percent(suffix = " %"),
      limits = c(0, ylim),
      expand = expansion()
    ),
    theme(
      panel.grid.minor.y = element_line(linetype = "dashed")
    )
  )
}
```

# Load data

Load polls.
  
```{r}
cal_raw <- read_parquet(config::get("calendar_long", config = "gold"))

head(cal_raw)
```

Filter calendar for our parties.

```{r}
INCLUDED_PARTIES <- c(
  "768", # ANO
  "1178", # Auto
  "1327", # SPOLU
  "166" # STAN
)

cal <-
  cal_raw |>
  filter(party %in% INCLUDED_PARTIES)
```

Load parties.

```{r}
parties <-
  read_parquet(config::get("parties", config = "silver")) |>
  filter(party %in% INCLUDED_PARTIES)

print(parties, n = 25)
```

Get only one poll per day.

```{r}
cal <-
  cal |>
  filter(!is.na(poll_id), !is.na(sample_size))

cal
```

Pivot wider.

```{r}
data_polls <-
  cal |>
  select(t, sample_size, agency_id, party, voteshare) |>
  pivot_wider(
    names_from = party,
    values_from = voteshare
  ) |>
  # Sort correctly
  select(t, sample_size, agency_id, all_of(INCLUDED_PARTIES))

data_polls
```

Create party mapping.

```{r}
party_map <-
  tibble(
    party = INCLUDED_PARTIES
  ) |>
  mutate(
    party_stan = row_number()
  ) |>
  left_join(parties, by = join_by(party)) |>
  bind_rows(tibble(
    party = "0000",
    party_stan = length(INCLUDED_PARTIES) + 1,
    party_shortcut = "X",
    party_name = "Ostatní",
    party_color = "#000"
  ))

party_map
```

Crate house mapping.

```{r}
house_map <-
  cal |>
  distinct(h = agency_id, agency) |>
  arrange(h) |>
  mutate(
    color = case_match(
      agency,
      "Ipsos" ~ "#002855", # Dark Blue
      "Kantar" ~ "#00A3E0", # Light Blue
      "Median" ~ "#005B9A", # Medium Blue
      "NMS Market Research" ~ "#93C01F", # Green
      "STEM" ~ "#DA251D", # Red
      .default = "#808080" # Fallback gray
    )
  )

house_map
```

Create poll-day-of-poll mapping.

```{r}
poll_days <-
  data_polls |>
  reframe(
    p = row_number(),
    t,
    h = agency_id
  )

poll_days
```

# Model

```{r run-models}
# Create&Save data
stan_data <- list(
  T = max(cal$t),
  P = nrow(data_polls),
  K = length(INCLUDED_PARTIES),
  H = max(data_polls$agency_id),
  y = as.matrix(data_polls[, INCLUDED_PARTIES]),
  yt = data_polls$t,
  yn = data_polls$sample_size,
  yh = data_polls$agency_id
)

saveRDS(
  object = stan_data,
  file = here::here("models", "model01", "input.RDS")
)

model <- cmdstan_model(
  stan_file = here::here("models", "model01", "model.stan")
)

# Create&Save prior model
stan_data$lik <- 0
fit_prior <- model$sample(
  data = stan_data,
  chains = 2,
  parallel_chains = 2,
)
fit_prior$save_object(
  file = here::here("models", "model01", "model_prior_fit.RDS")
)

# Create&Save posterior model
stan_data$lik <- 1
fit_post <- model$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4
)
fit_post$save_object(
  file = here::here("models", "model01", "model_posterior_fit.RDS")
)
```

Generate samples.

```{r fnc-pp_draws}
pp_draws <- function(prior_, posterior_, var, index_names = NULL) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    ) |>
    pivot_longer(
      cols = -c(.chain, .iteration, .draw, type),
      names_to = "variable"
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.numeric))
  }

  return(df)
}
```

```{r pp_draws}
samples_phi <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi",
  index_names = c("h")
)

samples_concentration <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "concentration",
  index_names = c("p")
)

samples_mu <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k")
)

samples_mu_raw_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_sigma",
  index_names = c("k")
)

samples_mu_raw_std <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_std",
  index_names = c("t", "k")
)

samples_lambda <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("p", "k")
)

samples_lambda_raw_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_sigma",
  index_names = c("k")
)

samples_lambda_raw_std <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_std",
  index_names = c("t", "k")
)

samples_y_hat <- pp_draws(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("p", "k")
)
```

Summarise samples.

```{r fnc-pp_summarise}
pp_summarise <- function(
  prior_,
  posterior_,
  var,
  index_names = NULL,
  statistics
) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$summary(variable = var, statistics) |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$summary(variable = var, statistics) |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.numeric))
  }

  return(df)
}
```

```{r pp_summarise}
summary_phi <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi",
  index_names = c("h"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_concentration <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "concentration",
  index_names = c("p"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu_raw_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("p", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda_raw_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_y_hat <- pp_summarise(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("p", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)
```

## Variable validation

### Phi

```{r}
summary_phi |>
  select(type, h, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, h)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = h, y = value, color = factor(h))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_houses(house_map) +
  labs(
    title = "Convergence statistics for parameter: phi",
    x = NULL,
    y = NULL
  )
```

```{r}
samples_phi |>
  select(type, h, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = h, y = value, color = factor(h)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  geom_hline(aes(yintercept = 1), linetype = "dashed", color = "lightblue") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_houses(house_map) +
  labs(
    title = "Copmarison of distributions for parameter: phi",
    x = NULL,
    y = NULL
  )
```

### Concentration

```{r}
summary_concentration |>
  select(type, p, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, p)
  ) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(h))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: concentration",
    x = NULL,
    y = NULL
  )
```

```{r}
summary_concentration |>
  select(type, p, mean, q5, q95) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = mean, color = factor(h))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(h)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_x_polls(cal = cal) +
  scale_color_houses(house_map) +
  labs(
    title = "Copmarison of distributions for parameter: concentration",
    x = NULL,
    y = NULL
  )
```

### Mu

```{r}
summary_mu |>
  select(type, t, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, t, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: mu",
    x = NULL,
    y = NULL
  )
```

```{r}
summary_mu |>
  select(type, t, k, mean, q5, q95) |>
  ggplot() +
  geom_line(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare() +
  labs(
    title = "Copmarison of distributions for parameter: mu",
    x = NULL,
    y = NULL
  )
```

### Mu raw sigma

```{r}
summary_mu_raw_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = k, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: mu raw sigma",
    x = NULL,
    y = NULL
  )
```

  
```{r}
samples_mu_raw_sigma |>
  select(type, k, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = k, y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: mu raw sigma",
    x = NULL,
    y = NULL
  )
```

### mu raw std

```{r}
samples_mu_raw_std |>
  ggplot() +
  geom_density(
    aes(x = value, color = factor(k), group = factor(t)),
    alpha = 0.2
  ) +
  facet_grid(rows = vars(k), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: mu raw std",
    x = NULL,
    y = NULL
  )
```

### Lambda

```{r}
summary_lambda |>
  select(type, p, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, p, k)
  ) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

```{r}
summary_lambda |>
  select(type, p, k, mean, q5, q95) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4,
    width = 10
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare() +
  labs(
    title = "Copmarison of distributions for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

### Lambda raw sigma

```{r}
summary_lambda_raw_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = k, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: lambda raw sigma",
    x = NULL,
    y = NULL
  )
```

```{r}
samples_lambda_raw_sigma |>
  select(type, k, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = k, y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: lambda raw sigma",
    x = NULL,
    y = NULL
  )
```

### mu raw std

```{r}
samples_lambda_raw_std |>
  ggplot() +
  geom_density(
    aes(x = value, color = factor(k), group = factor(t)),
    alpha = 0.2
  ) +
  facet_grid(rows = vars(k), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: lambda raw std",
    x = NULL,
    y = NULL
  )
```

## Posterior Predictive Check

```{r}
summary_mu_ <-
  summary_mu |>
  filter(k != 5, type == "posterior")

summary_lambda_ <-
  summary_lambda |>
  filter(k != 5, type == "posterior") |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one")

cal_ <-
  cal |>
  filter(t > 0) |>
  select(t, agency_id, voteshare, party) |>
  left_join(party_map, by = join_by(party), relationship = "many-to-one")
```

```{r}
ggplot() +
  # Plot estimated mu
  geom_line(
    data = summary_mu_,
    mapping = aes(x = t, y = mean, color = factor(k))
  ) +
  geom_errorbar(
    data = summary_mu_,
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.3
  ) +
  scale_color_parties(parties = party_map) +
  new_scale_color() +
  # Plot estimated lambda
  geom_point(
    data = summary_lambda_,
    mapping = aes(x = t, y = mean, color = factor(h)),
    shape = 4
  ) +
  geom_errorbar(
    data = summary_lambda_,
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(h)),
    alpha = 0.6,
    width = 5
  ) +
  # Plot true polls
  geom_point(
    data = cal_,
    mapping = aes(x = t, y = voteshare, color = factor(agency_id))
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_houses(houses = house_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare(ylim = 0.5) +
  labs(
    title = "Full posterior predictive checks",
    x = NULL,
    y = NULL
  )
```