---
title: Model 01
---

- Ignore partie mergers
- Ignore sample sizes
- Ignore more than one poll at one day
- Ignore errors

# Preparation

Load libraries.

```{r}
# Data
library(nanoparquet)
library(dplyr)
library(tidyr)
library(stringr)
# Bayes
library(cmdstanr)
library(posterior)
library(bayesplot)
# Ploting
library(ggplot2)
```

Check Stan installation.

```{r}
cmdstanr::check_cmdstan_toolchain()
```

ggplot.

```{r}
scale_color_parties <- function(parties) {
  scale_color_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$party_stan),
    labels = setNames(parties$party_shortcut, parties$party_stan)
  )
}

scale_fill_parties <- function(parties) {
  scale_fill_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$party_stan),
    labels = setNames(parties$party_shortcut, parties$party_stan)
  )
}

scale_x_parties <- function(parties) {
  list(
    scale_x_continuous(
      breaks = parties$party_stan,
      labels = parties$party_shortcut
    ),
    theme(
      panel.grid.major.x = element_blank()
    )
  )
}

scale_x_polls <- function(cal) {
  list(
    scale_x_continuous(
      name = NULL,
      breaks = seq(from = 0, to = max(cal$t), by = 25),
      minor_breaks = seq(from = 0, to = max(cal$t), by = 25/2),
      limits = c(0, max(cal$t)),
      expand = expansion(add = 5)
    ),
    theme(
      panel.grid.minor.x = element_line(linetype = "dashed")
    )
  )
}

scale_y_voteshare <- function(ylim = 1) {
  list(
    scale_y_continuous(
      name = "voteshare [%]",
      breaks = seq(0, 1, by = 0.25),
      minor_breaks = seq(0, 1, by = 0.25/2),
      labels = scales::label_percent(suffix = " %"),
      limits = c(0, ylim),
      expand = expansion()
    ),
    theme(
      panel.grid.minor.y = element_line(linetype = "dashed")
    )
  )
}
```

# Load data

Load polls.
  
```{r}
cal <- read_parquet(config::get("calendar_long", config = "gold"))

head(cal)
```

Filter calendar for our parties.

```{r}
no_merge_parties <- c(
  "768", # ANO
  "1178", # Auto
  "1327", # SPOLU
  "166" # STAN
)

cal <-
  cal |>
  filter(party %in% no_merge_parties)
```

Load parties.

```{r}
parties <-
  read_parquet(config::get("parties", config = "silver")) |>
  filter(party %in% cal$party)

print(parties, n = 25)
```

Get only one poll per day.

```{r}
df_full <-
  cal |>
  filter(!is.na(poll_id), !is.na(sample_size)) |>
  slice_max(poll_id, by = t)

df_full
```

Prepare data.

```{r}
data_polls <-
  df_full |>
  filter(party %in% no_merge_parties) |>
  select(t, party, voteshare, sample_size) |>
  pivot_wider(
    names_from = party,
    values_from = voteshare
  )

data_polls
```

Extract initial state.

```{r}
initial_poll <-
  data_polls |>
  filter(t == 0) |>
  select(all_of(no_merge_parties)) |>
  unlist()

print(initial_poll)

data_polls <-
  data_polls |>
  filter(t != 0)

data_polls
```

Map parties to names.

```{r}
party_map <-
  data_polls |>
  select(all_of(no_merge_parties)) |>
  pivot_longer(cols = everything()) |>
  distinct(party = name) |>
  mutate(
    party_stan = row_number()
  ) |>
  left_join(parties, by = join_by(party))

party_map
```

Make sure the data is correctly ordered.

```{r}
data_polls <-
  data_polls |>
  select(t, sample_size, all_of(party_map$party))

data_polls
```

# Model

```{r}
stan_data <- list(
  T = max(cal$t),
  P = nrow(data_polls),
  K = length(no_merge_parties),
  y = as.matrix(data_polls[, no_merge_parties]),
  yt = data_polls$t,
  yn = data_polls$sample_size,
  initial_state = initial_poll
)

saveRDS(
  object = stan_data,
  file = here::here("models", "model01", "input.RDS")
)

model <- cmdstan_model(
  stan_file = here::here("models", "model01", "model.stan")
)

stan_data$lik <- 0
fit_prior <- model$sample(
  data = stan_data,
  chains = 1
)
fit_prior$save_object(
  file = here::here("models", "model01", "model_prior_fit.RDS")
)

stan_data$lik <- 1
fit_post <- model$sample(
  data = stan_data,
  chains = 1
)
fit_post$save_object(
  file = here::here("models", "model01", "model_posterior_fit.RDS")
)
```

Add Other category.

```{r}
# Add "Others"
party_map <- bind_rows(party_map, tibble(
  party = "0000",
  party_stan = max(party_map$party_stan) + 1,
  party_shortcut = "X",
  party_name = "Ostatní",
  party_color = "#000"
))
```

Generate samples.

```{r}
pp_draws <- function(prior_, posterior_, var, index_names = NULL) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    ) |>
    pivot_longer(
      cols = -c(.chain, .iteration, .draw, type),
      names_to = "variable"
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.numeric))
  }

  return(df)
}
```

```{r}
samples_phi <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi"
)

samples_mu <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k")
)

samples_mu_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_sigma",
  index_names = c("k")
)

samples_lambda <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("t", "k")
)

samples_lambda_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_sigma",
  index_names = c("k")
)

samples_y_hat <- pp_draws(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("t", "k")
)
```

Summarise samples.

```{r}
pp_summarise <- function(
  prior_,
  posterior_,
  var,
  index_names = NULL,
  statistics
) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$summary(variable = var, statistics) |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$summary(variable = var, statistics) |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.numeric))
  }

  return(df)
}
```

```{r}
summary_phi <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi",
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("t", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_y_hat <- pp_summarise(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("t", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)
```

## Variable validation

### Phi

```{r}
summary_phi |>
  select(type, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = 1, y = value)) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  labs(
    title = "Convergence statistics for parameter: phi",
    x = NULL,
    y = NULL
  )
```

```{r}
samples_phi |>
  select(type, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = 1, y = value),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  labs(
    title = "Copmarison of distributions for parameter: phi",
    x = NULL,
    y = NULL
  )
```

### Mu

```{r}
summary_mu |>
  select(type, t, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, t, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: mu",
    x = NULL,
    y = NULL
  )
```

```{r}
summary_mu |>
  select(type, t, k, mean, q5, q95) |>
  ggplot() +
  geom_line(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Copmarison of distributions for parameter: mu",
    x = NULL,
    y = NULL
  )
```

### Mu sigma

```{r}
summary_mu_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = k, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: mu sigma",
    x = NULL,
    y = NULL
  )
```

  
```{r}
samples_mu_sigma |>
  select(type, k, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = k, y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: mu sigma",
    x = NULL,
    y = NULL
  )
```

### Lambda

```{r}
summary_lambda |>
  select(type, t, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, t, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

```{r}
summary_lambda |>
  select(type, t, k, mean, q5, q95) |>
  ggplot() +
  geom_line(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare() +
  labs(
    title = "Copmarison of distributions for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

### Lambda sigma

```{r}
summary_lambda_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = k, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: lambda sigma",
    x = NULL,
    y = NULL
  )
```
  
```{r}
samples_lambda_sigma |>
  select(type, k, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = k, y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: lambda sigma",
    x = NULL,
    y = NULL
  )
```

## Posterior Predictive Check

```{r}
known_polls <- 
  cal |> 
  filter(t > 0) |> 
  select(t, voteshare, party) |> 
  left_join(party_map, by = join_by(party))
```

```{r}
summary_lambda |> 
  select(type, t, k, mean, q5, q95) |> 
  filter(k != 5) |>
  ggplot() +
  geom_line(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  geom_point(
    data = known_polls,
    mapping = aes(x = t, y = voteshare, color = factor(party_stan))
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare() +
  labs(
    title = "Posterior predictive check for: lambda",
    x = NULL,
    y = NULL
  )
```

```{r}
known_summary_y_hat <-
  summary_y_hat |> 
  filter(k != 5)

ggplot() +
  geom_line(
    data = known_summary_y_hat,
    mapping = aes(x = t, y = mean, color = factor(k))
  ) +
  geom_errorbar(
    data = known_summary_y_hat,
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  geom_point(
    data = known_polls,
    mapping = aes(x = t, y = voteshare, color = factor(party_stan))
  ) +
  theme_bw() +
  scale_color_parties(party_map) +
  scale_y_voteshare(ylim = 1) +
  labs(
    title = "Posterior predictive check for: y_hat",
    x = NULL,
    y = NULL
  )
```

```{r}

```