---
title: Model 01 - Only complete parties
author: Michal Lauer
date: today
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
fig-width: 10
fig-height: 8
---

# Introduction

This project analyses the underlying party support measured through polls conducted by various agencies.
Full code can be found on [GitHub](https://github.com/MichalLauer/Elections-2025).

## Data

The data in this analysis are scraped from [programydovoleb.cz](https://programydovoleb.cz/) and contain all
polls that have been conducted in regards to the elections to the House of Representatives, 2025 (Volby do poslanecké
sněmovny v r. 2025). The model does not work with the final results.

## Limitations

- Only parties with no missing data are included in the analysis.
- No hierarchical priors are currently set.
- Polls with missing sample size are excluded.
- Party mergers are ignored (e.g. Piráti + Zelení).
- It is assumed that all parties are independent, a rather *naïve* assumption.
- The *Others* party (X) is not usually interpreted as it is a mix of all other parties, thus it would not make much sense.
- Since this model contains only 4 parties, comparison to the final results should be made with caution.

# Preparation

Load libraries.

```{r packages, output=FALSE}
# Data
library(nanoparquet)
library(dplyr)
library(tidyr)
library(stringr)
# Bayes
library(cmdstanr)
library(posterior)
library(bayesplot)
# Ploting
library(ggplot2)
library(ggnewscale)
```

Check Stan installation.

```{r stan-check}
options("mc.cores" = future::availableCores() - 1)
cmdstanr::check_cmdstan_toolchain()
```

ggplot2 helpers.

```{r ggplot-utils}
scale_color_parties <- function(parties) {
  scale_color_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$k),
    labels = setNames(parties$party_shortcut, parties$k)
  )
}

scale_color_houses <- function(houses) {
  list(
    scale_color_manual(
      name = "Polling House",
      breaks = houses$h,
      values = setNames(houses$color, houses$h),
      labels = setNames(houses$agency, houses$h)
    ),
    theme(
      legend.position = "bottom"
    )
  )
}

scale_fill_parties <- function(parties) {
  scale_fill_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$k),
    labels = setNames(parties$party_shortcut, parties$k)
  )
}

scale_x_parties <- function(parties) {
  list(
    scale_x_discrete(
      breaks = parties$k,
      labels = parties$party_shortcut
    ),
    theme(
      panel.grid.major.x = element_blank()
    )
  )
}

scale_x_polls <- function(cal) {
  list(
    scale_x_continuous(
      name = "Days since first poll",
      breaks = seq(from = 1, to = max(cal$t), by = 50),
      minor_breaks = seq(from = 0, to = max(cal$t), by = 25),
      limits = c(0, max(cal$t)),
      expand = expansion(add = 5)
    ),
    theme(
      panel.grid.minor.x = element_line(linetype = "dashed")
    )
  )
}

scale_x_houses <- function(houses) {
  list(
    scale_x_discrete(
      breaks = houses$h,
      labels = houses$agency
    ),
    theme(
      panel.grid.major.x = element_blank()
    )
  )
}

scale_y_voteshare <- function(ylim = 1) {
  list(
    scale_y_continuous(
      name = "voteshare [%]",
      breaks = seq(0, 1, by = 0.05),
      minor_breaks = seq(0, 1, by = 0.01),
      labels = scales::label_percent(suffix = " %"),
      limits = c(0, ylim),
      expand = expansion(),
      guide = guide_axis(minor.ticks = TRUE)
    ),
    guides(
      y.sec = guide_axis(minor.ticks = TRUE, title = "voteshare [%]")
    ),
    theme(
      panel.grid.minor.y = element_line(linetype = "dashed"),
      axis.ticks.y.right = element_line(),
      axis.minor.ticks.y.right = element_line(),
      axis.minor.ticks.length = unit(-0.1, "cm")
    )
  )
}
```

# Load data

Load polls.
  
```{r}
cal_raw <- read_parquet(config::get("calendar_long", config = "gold"))

head(cal_raw)
```

Filter calendar for our parties.

```{r}
INCLUDED_PARTIES <- c(
  "768", # ANO
  "1178", # Auto
  "1327", # SPOLU
  "166" # STAN
)

cal <-
  cal_raw |>
  filter(party %in% INCLUDED_PARTIES)
```

Load parties.

```{r}
parties <-
  read_parquet(config::get("parties", config = "silver")) |>
  filter(party %in% INCLUDED_PARTIES)

parties
```

Get only complete polls.

```{r}
cal <-
  cal |>
  filter(!is.na(sample_size))

cal
```

Pivot wider.

```{r}
data_polls <-
  cal |>
  select(t, sample_size, agency_id, party, voteshare) |>
  pivot_wider(
    names_from = party,
    values_from = voteshare
  ) |>
  # Sort correctly
  select(t, sample_size, agency_id, all_of(INCLUDED_PARTIES))

data_polls
```

Create party mapping.

```{r}
party_map <-
  tibble(
    party = INCLUDED_PARTIES
  ) |>
  mutate(
    k = row_number()
  ) |>
  left_join(parties, by = join_by(party)) |>
  bind_rows(tibble(
    party = "0000",
    k = length(INCLUDED_PARTIES) + 1,
    party_shortcut = "X",
    party_name = "Ostatní",
    party_color = "#000"
  ))

party_map
```

Crate house mapping.

```{r}
house_map <-
  cal |>
  distinct(h = agency_id, agency = as.character(agency)) |>
  arrange(h) |>
  mutate(
    color = recode_values(
      agency,
      "Ipsos" ~ "#002855",
      "Kantar" ~ "#00A3E0",
      "Median" ~ "#005B9A",
      "NMS Market Research" ~ "#93C01F",
      "STEM" ~ "#DA251D",
      default = "#808080"
    )
  )

house_map
```

Create poll-day-agency mapping.

```{r}
poll_days <-
  data_polls |>
  reframe(
    p = row_number(),
    t,
    h = agency_id
  )

poll_days
```

# Model

$$
\begin{aligned}
y_p &\sim Dir(\lambda_p * \kappa_p) \\
\kappa_p &= \phi_h \cdot n_p \\
\log(\phi_h) &\sim N(0, 1) \\
\text{CLR(}\lambda_p\text{)} &\sim N(\mu_p + \beta_h, \sigma_\lambda) \\
\sigma_\lambda &\sim \text{Exponential}(80) \\
\text{CLR(}\mu_1\text{)} &\sim N(\text{CLR(}y_1\text{)}, \sigma_\mu) \\
\text{CLR(}\mu_t\text{)} &\sim N(\text{CLR(}\mu_{t-1}\text{)}, \sigma_\mu) \\
\sigma_\mu &\sim \text{Exponential}(80) \\
\sigma_\lambda &\sim \text{Exponential}(80) \\
\beta_h &\sim N\left(0,\ 0.1 * \sqrt{(K+1)/K}\right) \\
\sum_{h=1}^H \beta_h &= 0 \\
\end{aligned}
$$

where

- $y_p = [y_{p,1}, \ldots, y_{p,K + 1}]'$ - observed voteshare of poll $p = 1, \dots, P$ for parties $1, \dots, K + 1$,
- $\kappa_p$ - effective sample size of poll $p$,
- $\phi_h$ - precision of house $h$,
- $n_p$ - reported sample size of poll $p$,
- $\text{CLR(}\lambda_p\text{)}$ - centered log-ratio transformed latent parameter $\lambda_p$,
  - $\lambda_p$ - latent variable in simplex space,
- $\sigma_\lambda = [\sigma_{\lambda,1}, \ldots, \sigma_{\lambda,K + 1}]'$ - standard deviations of the latent variables $\lambda_k$ for $k = 1, \dots, K + 1$,
- $\text{CLR(}\mu_t\text{)} = [\mu_{t,1}, \ldots, \mu_{t,K + 1}]'$ - true underlying support of party $k = 1, \dots, K$ at time $t = 1, \dots, T$ in CLR space,
  - $\mu_t$ - latent variable in simplex space,
- $\sigma_\mu = [\sigma_{\mu,1}, \ldots, \sigma_{\mu,K + 1}]'$ - standard deviations of the latent variables $\mu_k$ for $k = 1, \dots, K + 1$.
- $\beta_h = [\beta_{h,1}, \ldots, \beta_{h,K + 1}]'$ - house-specific biases for party $k = 1, \dots, K + 1$ in house $h = 1, \dots, H$.

Notes:

- The $K + 1$-th party is computed dynamically in Stan and it is the "Others" party.

```{r run-models}
# Create&Save data
stan_data <- list(
  T = max(cal$t),
  P = nrow(data_polls),
  K = length(INCLUDED_PARTIES),
  H = max(data_polls$agency_id),
  y = as.matrix(data_polls[, INCLUDED_PARTIES]),
  yt = data_polls$t,
  yn = data_polls$sample_size,
  yh = data_polls$agency_id
)

saveRDS(
  object = stan_data,
  file = here::here("models", "model01", "input.RDS")
)

model <- cmdstan_model(
  stan_file = here::here("models", "model01", "model.stan")
)

# Create&Save prior model
stan_data$lik <- 0
fit_prior <- model$sample(
  data = stan_data,
  chains = 2,
  parallel_chains = 2,
)
fit_prior$save_object(
  file = here::here("models", "model01", "model_prior_fit.RDS")
)

# Create&Save posterior model
stan_data$lik <- 1
fit_post <- model$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4
)
fit_post$save_object(
  file = here::here("models", "model01", "model_posterior_fit.RDS")
)
```

Generate samples.

```{r fnc-pp_draws}
pp_draws <- function(prior_, posterior_, var, index_names = NULL) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$draws(variables = var) |>
      as_draws_df() |>
      as_tibble() |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    ) |>
    pivot_longer(
      cols = -c(.chain, .iteration, .draw, type),
      names_to = "variable"
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.integer))
  }

  return(df)
}
```

```{r pp_draws}
samples_phi <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi",
  index_names = c("h")
)

samples_kappa <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "kappa",
  index_names = c("p")
)

samples_house_bias <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "house_bias",
  index_names = c("h", "k")
)

samples_mu <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k")
)

samples_mu_raw_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_sigma",
  index_names = c("k")
)

samples_mu_raw_std <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_std",
  index_names = c("t", "k")
)

samples_lambda <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("p", "k")
)

samples_lambda_raw_sigma <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_sigma",
  index_names = c("k")
)

samples_lambda_raw_std <- pp_draws(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_std",
  index_names = c("t", "k")
)

samples_y_hat <- pp_draws(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("p", "k")
)
```

Summarise samples.

```{r fnc-pp_summarise}
pp_summarise <- function(
  prior_,
  posterior_,
  var,
  index_names = NULL,
  statistics
) {
  df_prior <- NULL
  if (!is.null(prior_)) {
    df_prior <-
      prior_$summary(variable = var, statistics) |>
      mutate(type = "prior")
  }

  df_posterior <- NULL
  if (!is.null(posterior_)) {
    df_posterior <-
      posterior_$summary(variable = var, statistics) |>
      mutate(type = "posterior")
  }

  df <-
    bind_rows(df_prior, df_posterior) |>
    mutate(
      type = ordered(type, levels = c("prior", "posterior"))
    )

  if (length(index_names) > 0) {
    df <-
      df |>
      mutate(indices = str_extract(variable, r"(\[(.+)\])", group = 1)) |>
      separate_wider_delim(
        cols = indices,
        delim = ",",
        names = index_names
      ) |>
      mutate(across(.cols = all_of(index_names), .fns = as.integer))
  }

  return(df)
}
```

```{r pp_summarise}
summary_phi <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "phi",
  index_names = c("h"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_kappa <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "kappa",
  index_names = c("p"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_house_bias <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "house_bias",
  index_names = c("h", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu",
  index_names = c("t", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_mu_raw_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "mu_raw_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda",
  index_names = c("p", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_lambda_raw_sigma <- pp_summarise(
  prior_ = fit_prior,
  posterior_ = fit_post,
  var = "lambda_raw_sigma",
  index_names = c("k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)

summary_y_hat <- pp_summarise(
  prior_ = NULL,
  posterior_ = fit_post,
  var = "y_hat",
  index_names = c("p", "k"),
  statistics = c(default_convergence_measures(), default_summary_measures())
)
```

## Variable validation

### Phi

```{r}
summary_phi |>
  select(type, h, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, h)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = factor(h), y = value, color = factor(h))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_houses(house_map) +
  labs(
    title = "Convergence statistics for parameter: phi",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

```{r}
samples_phi |>
  select(type, h, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = factor(h), y = value, color = factor(h)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  geom_hline(aes(yintercept = 1), linetype = "dashed", color = "lightblue") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_houses(house_map) +
  labs(
    title = "Copmarison of distributions for parameter: phi",
    x = NULL,
    y = NULL
  )
```

The prior assumes that the phi parameter (*sample_size* multiplier) is
centered around 1 with 95% credible intervals around 0.6 and 1.6. We are
saying implying that every pollster can have variously effective sample sizes. Their 'strength' can be smaller by 40 \% or larger by 60 \%.

```{r}
samples_phi |>
  group_by(type, variable) |>
  summarise(
    mean = mean(value),
    sd = sd(value),
    q5 = quantile(value, 0.05),
    q95 = quantile(value, 0.95),
    `Pr(x > 1)` = mean(value > 1),
    `Pr(x < 1)` = mean(value < 1),
  )
```

After fitting the model, it can be observed that Ipsos, Kantar and STEM have most probably stronger sample sizes
compared to other pollsters. This means that their population weights and sampling estimates are likely more reliable.
Pollster Median is probably the least reliable. NMS Market Research has moved to an average of around 0,95 and while 
the posterior is certainly less variable, it's central tendency has not moved much.

### Kappa

```{r}
summary_kappa |>
  select(type, p, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, p)
  ) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(h))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: concentration",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK. Even though they all look the same, this is just a scaling issue, as
kappa for every pollster is dependent on the actual sample size.

```{r}
samples_kappa |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  left_join(house_map, by = join_by(h), relationship = "many-to-one") |>
  group_by(type, agency) |>
  summarise(
    mean = mean(value),
    sd = sd(value)
  )

```

The posterior effective sample size is larger for pollsters with phi > 1. Even though three polsters (Kantar, Median,
NMS) have very simmilar starting prior effective sample size, the posterior is different for every one of them. This
suggests that the model is able to identify pollsters whose polls estimate better the underlying support.

```{r}
summary_kappa |>
  select(type, p, mean, q5, q95) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = mean, color = factor(h))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(h)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_x_polls(cal = cal) +
  scale_color_houses(house_map) +
  labs(
    title = "Copmarison of distributions for parameter: concentration",
    x = NULL,
    y = NULL
  )
```

### House bias

```{r}
summary_house_bias |>
  select(type, h, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, h, k)
  ) |>
  left_join(house_map, by = join_by(h), relationship = "many-to-one") |>
  left_join(party_map, by = join_by(k), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = factor(h), y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(party_map) +
  scale_x_houses(house_map) +
  labs(
    title = "Convergence statistics for parameter: house bias",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

```{r}
samples_house_bias |>
  select(type, h, k, value) |>
  filter(k != 5) |>
  left_join(house_map, by = join_by(h), relationship = "many-to-one") |>
  left_join(party_map, by = join_by(k), relationship = "many-to-one") |>
  ggplot() +
  geom_violin(
    mapping = aes(x = factor(k), y = value, color = factor(h)),
    position = position_dodge(width = 0.8)
  ) +
  geom_vline(
    data = tibble(xint = seq(1.5, 3.5, 1)),
    aes(xintercept = xint),
    linetype = "dashed"
  ) +
  geom_hline(aes(yintercept = 1), linetype = "dashed", color = "lightblue") +
  facet_wrap(vars(type), ncol = 1) +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_parties(party_map) +
  labs(
    title = "Convergence statistics for parameter: house bias",
    x = NULL,
    y = NULL
  )
```

The prior bias for every party of every polster is identical, centered around 1, typically ranging from 0.8 to 1.2.
This implies that the typical pollster bias is around 20 % (beta represents odds-ratio). The posterior estimates
suggest that 3/5 pollsters typically underestimated ANO by around 5 - 15 % (for example, they would estimate support
of 35 % * 0,85 = 29,75 %). AUTO party was overestimated by 3/5 pollsters by around 5 % and SPOLU was slightly
underestimated by around 2 %. The posterior suggest that STAN was underestimated the most and no pollsters has
overstimated them. The underestimation is by around 10 % from 3/5 pollsters.

### Mu

```{r}
summary_mu |>
  select(type, t, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, t, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: mu",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

::: {.callout-caution collapse="true"}
It is interesting that the prior and posterior statistics look like a *chain* - why?
:::

```{r}
summary_mu |>
  select(type, t, k, mean, q5, q95) |>
  filter(k != 5) |>
  ggplot() +
  geom_line(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare(ylim = 0.6) +
  labs(
    title = "Comparison of distributions for parameter: mu",
    x = NULL,
    y = NULL
  )
```

The prior variance of every party grows in time, because the process is assumed be a random walk. However, even after 
around a year of not observing a single poll, the party support of the strongest party most probably won't grow over
50 %. Without observing any other polls, the underlying support also does not change and is constant, with increasing
variance.

::: {.callout-caution collapse="true"}
Theoretically, the variance should not be the same and for larger parties, it should be smaller compared to smaller
parties. Also, for AUTO, the variance is not symmetrical as it could grow belo 0 and it is thus cut. Needs to be
dealt with.
:::

The posterior shows the estimated support for each party.

::: {.callout-caution collapse="true"}
The variance should be much smaller on days when the poll is observed, and should grow the longer we have without a 
poll. I *think* this is happening, because at the start, this can be observed. Later on, there is many polls and thus,
this effect is probably not seen. Need to check.
:::

### Mu raw sigma

```{r}
summary_mu_raw_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = factor(k), y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: mu raw sigma",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

```{r}
samples_mu_raw_sigma |>
  select(type, k, value) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = factor(k), y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: mu raw sigma",
    x = NULL,
    y = NULL
  )
```

Representation of variance on the log scale, and thus, it is not very interpretable. The posterior is however largest
for AUTO, implying that their support has been the most shakiest.

### mu raw std

```{r}
samples_mu_raw_std |>
  ggplot() +
  geom_density(
    aes(x = value, color = factor(k), group = factor(t)),
    alpha = 0.2
  ) +
  facet_grid(rows = vars(k), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: mu raw std",
    x = NULL,
    y = NULL
  )
```

This is the representation of non-centered latent parameter mu. All these samples should theoretically have standard
normal distribution. If not, it could indicate a miss-specified model where some of the variance in the latent
variable is hidden in this non-centered variable. There are no large deviations from the expected distribution.

### Lambda

```{r}
summary_lambda |>
  select(type, p, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, p, k)
  ) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  labs(
    title = "Convergence statistics for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

```{r}
summary_lambda |>
  select(type, p, k, mean, q5, q95) |>
  filter(k != 5) |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one") |>
  ggplot() +
  geom_point(mapping = aes(x = t, y = mean, color = factor(k))) +
  geom_errorbar(
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.4,
    width = 10
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare(ylim = 0.6) +
  labs(
    title = "Comparison of distributions for parameter: lambda",
    x = NULL,
    y = NULL
  )
```

Lambda represents individual polls and their variance. Again, it can be observed (apriori) that the longer the model
goes without new polls, the larger the variance is. The underlying support measured by the poll is again not changing.
When the model accounts for data, the posterior intervals are much smaller.

### Lambda raw sigma

```{r}
summary_lambda_raw_sigma |>
  select(type, k, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = -c(type, k)
  ) |>
  ggplot() +
  geom_point(mapping = aes(x = factor(k), y = value, color = factor(k))) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: lambda raw sigma",
    x = NULL,
    y = NULL
  )
```

Prior and posterior diagnostics are OK.

```{r}
samples_lambda_raw_sigma |>
  select(type, k, value) |>
  filter(k != 5) |>
  ggplot() +
  geom_violin(
    mapping = aes(x = factor(k), y = value, fill = factor(k)),
    alpha = 0.4
  ) +
  facet_grid(cols = vars(type), scales = "free") +
  theme_bw() +
  scale_fill_parties(parties = party_map) +
  scale_x_parties(parties = party_map) +
  labs(
    title = "Copmarison of distributions for parameter: lambda raw sigma",
    x = NULL,
    y = NULL
  )
```

The variance is defined in log odds and is thus not clearly interpretable. The posterior distributions do not change
much for ANO, Spolu and STAN - they became primarly narrower. For AUTO, the variance is much larger and the tail
is very large, suggesting (again) that the variance of party is inconsistent.

### lambda raw std

```{r}
samples_lambda_raw_std |>
  ggplot() +
  geom_density(
    aes(x = value, color = factor(k), group = factor(t)),
    alpha = 0.2
  ) +
  facet_grid(rows = vars(k), cols = vars(type), scales = "free") +
  theme_bw() +
  scale_color_parties(parties = party_map) +
  labs(
    title = "Convergence statistics for parameter: lambda raw std",
    x = NULL,
    y = NULL
  )
```

This is the representation of non-centered latent parameter lambda. All these samples should theoretically have standard
normal distribution. If not, it could indicate a miss-specified model where some of the variance in the latent
variable is hidden in this non-centered variable. There are no large deviations from the expected distribution.

## Posterior Predictive Check

Prepare data.

```{r}
summary_mu_ <-
  summary_mu |>
  filter(k != 5, type == "posterior")

summary_lambda_ <-
  summary_lambda |>
  filter(k != 5, type == "posterior") |>
  left_join(poll_days, by = join_by(p), relationship = "many-to-one")

cal_ <-
  cal |>
  filter(t > 0) |>
  select(t, agency_id, voteshare, party) |>
  left_join(party_map, by = join_by(party), relationship = "many-to-one")
```

Plot PPC.

```{r}
ggplot() +
  # Plot estimated mu
  geom_line(
    data = summary_mu_,
    mapping = aes(x = t, y = mean, color = factor(k))
  ) +
  geom_errorbar(
    data = summary_mu_,
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(k)),
    alpha = 0.3
  ) +
  scale_color_parties(parties = party_map) +
  new_scale_color() +
  # Plot estimated lambda
  geom_point(
    data = summary_lambda_,
    mapping = aes(x = t, y = mean, color = factor(h)),
    shape = 4
  ) +
  geom_errorbar(
    data = summary_lambda_,
    mapping = aes(x = t, ymin = q5, ymax = q95, color = factor(h)),
    alpha = 0.6,
    width = 5
  ) +
  # Plot true polls
  geom_point(
    data = cal_,
    mapping = aes(x = t, y = voteshare, color = factor(agency_id))
  ) +
  theme_bw() +
  scale_color_houses(house_map) +
  scale_x_polls(cal = cal) +
  scale_y_voteshare(ylim = 0.5) +
  labs(
    title = "Full posterior predictive checks",
    x = NULL,
    y = NULL
  )
```

The line, colored by each party, represents the true underlying party support measured by mu. This latent variable
is defined at every time point t and it's variance represents the uncertainty about the true support for each party.
The dots, colored by polling houses, represent the observed party support through the pollsters poll. They are
defined only on days when there was at least one poll. This observed data point does not have any variance nor uncertainty displayed. The cross, colored by the same way as observed polls, represents the average estimated
poll support with 95 % credible intervals. This is the posterior distribution of party support from a given poll.

Look at the second and third poll for party STAN. These polls happened at the same day, and are thus displayed on
the same x-axis line. The blue dot - poll conducted by kantar - is much closer to the estimated true party 
support. It is also very close to the estimated posterior average poll support, the latent variable lambda. On the
other hand, the red dot - poll by STEM - is much lower and lies outside of the estimated credible intervals for both
the underlying support mu and the poll uncertainty lambda. This suggest that this poll is an outlier - for some reason,
STEM estimated the support to be much lower than it probably was, and the model is able to measure this discrepancy.
This can be observed on some polls for all parties.

Finally, the estimated party support as a whole can be compared to the poll estimates. The ANO party has been
consistently underestimated by STEM and NMS, and most of the polls conducted are way outside of the 95%
credible intervals. On the other hand, party ODS has been consistently overestimated by all pollsters, with most
polls falling well above the credible intervals. On the other hand, the SPOLU party was, on average, estaimted
well by the polls.