---
title: Model 01
---

- Ignore partie mergers
- Ignore sample sizes
- Ignore more than one poll at one day
- Ignore errors

# Preparation

Load libraries.

```{r}
# Data
library(nanoparquet)
library(dplyr)
library(tidyr)
library(stringr)
# Bayes
library(cmdstanr)
library(posterior)
library(bayesplot)
# Ploting
library(ggplot2)
```

Check Stan installation.

```{r}
cmdstanr::check_cmdstan_toolchain()
```

ggplot.

```{r}
scale_color_parties <- function(parties) {
  scale_color_manual(
    name = "Party",
    values = setNames(parties$party_color, parties$party_shortcut)
  )
}

scale_x_polls <- function(cal) {
  scale_x_continuous(
    name = NULL,
    breaks = seq(from = 0, to = max(cal$t), by = 25),
    limits = c(0, max(cal$t)),
    expand = expansion(add = 5)
  )
}

scale_x_parties <- function(party_map) {
  scale_x_continuous(
    breaks = party_map$party_stan,
    labels = party_map$party_shortcut
  )
}

scale_shape_type <- function() {
  scale_shape_manual(
    name = "Model type",
    values = c("prior" = 19, "posterior" = 4),
    breaks = c("prior", "posterior")
  )
}

scale_color_type <- function() {
  scale_color_manual(
    name = "Model type",
    values = c("prior" = "lightgreen", "posterior" = "darkblue"),
    breaks = c("prior", "posterior")
  )
}

scale_fill_type <- function() {
  scale_fill_manual(
    name = "Model type",
    values = c("prior" = "lightgreen", "posterior" = "darkblue"),
    breaks = c("prior", "posterior")
  )
}

scale_y_voteshare <- function() {
  scale_y_continuous(
    name = "voteshare [%]",
    labels = scales::label_percent(suffix = " %")
  )
}
```

# Load data

Load polls.
  
```{r}
cal <- read_parquet(config::get("calendar_long", config = "gold"))

head(cal)
```

Filter calendar for our parties.

```{r}
no_merge_parties <- c(
  "768", # ANO
  "1178", # Auto
  "1327", # SPOLU
  "166" # STAN
)

cal <-
  cal |>
  filter(party %in% no_merge_parties)
```

Load parties.

```{r}
parties <-
  read_parquet(config::get("parties", config = "silver")) |>
  filter(party %in% cal$party)

print(parties, n = 25)
```

Get only one poll per day.

```{r}
df_full <-
  cal |>
  filter(!is.na(poll_id)) |>
  slice_max(poll_id, by = t)

df_full
```

Prepare data.

```{r}
data_polls <-
  df_full |>
  filter(party %in% no_merge_parties) |>
  select(t, party, voteshare) |>
  pivot_wider(
    names_from = party,
    values_from = voteshare
  )

data_polls
```

Extract initial state.

```{r}
initial_poll <-
  data_polls |>
  filter(t == 0) |>
  select(-t) |>
  unlist()

print(initial_poll)

data_polls <-
  data_polls |>
  filter(t != 0)

data_polls
```

Map parties to names.

```{r}
party_map <-
  data_polls |>
  select(-t) |>
  pivot_longer(cols = everything()) |>
  distinct(party = name) |>
  mutate(
    party_stan = row_number()
  ) |>
  left_join(parties, by = join_by(party))


party_map
```

Make sure the data is correctly ordered.

```{r}
data_polls <-
  data_polls |>
  select(t, all_of(party_map$party))

data_polls
```

# Model

```{r}
stan_data <- list(
  T = max(cal$t),
  P = nrow(data_polls),
  K = ncol(data_polls) - 1,
  y = as.matrix(data_polls[, -1]),
  yt = data_polls$t,
  initial_state = initial_poll
)

model <- cmdstan_model(
  stan_file = here::here("data", "0model", "model01", "model.stan")
)

stan_data$lik <- 0
fit_prior <- model$sample(
  data = stan_data,
  chains = 1
)

stan_data$lik <- 1
fit_post <- model$sample(
  data = stan_data,
  chains = 1
)
```

## Model Convergence

```{r}
# mcmc_trace(alpha_log)
```

# Variables

Generate samples.

```{r}
mu_raw_pri <- fit_prior$draws(variables = "mu_raw")
mu_raw_post <- fit_post$draws(variables = "mu_raw")

mu_pri <- fit_prior$draws(variables = "mu")
mu_post <- fit_post$draws(variables = "mu")

beta_pri <- fit_prior$draws(variables = "beta")
beta_post <- fit_post$draws(variables = "beta")

sigma_pri <- fit_prior$draws(variables = "sigma")

alpha_raw_pri <- fit_prior$draws(variables = "alpha_raw")
alpha_raw_post <- fit_post$draws(variables = "alpha_raw")

alpha_pri <- fit_prior$draws(variables = "alpha")
alpha_post <- fit_post$draws(variables = "alpha")
```

Summarise samples

```{r}
summarise_samples <- function(x, party, extract = "tk") {
  df <- x |>
    summarise_draws(
      mean,
      "sd",
      ~ quantile(.x, probs = c(0.055, 0.945)),
      default_convergence_measures()
    )

  if (!is.null(extract)) {
    if (extract == "t") {
      df <-
        df |>
        mutate(
          t = as.numeric(str_extract(variable, r"(\[(\d+)])", group = 1))
        )
    } else if (extract == "k") {
      df <-
        df |>
        mutate(
          k = as.numeric(str_extract(variable, r"(\[(\d+)\])", group = 1))
        )
    } else if (extract == "tk") {
      df <-
        df |>
        mutate(
          t = as.numeric(str_extract(variable, r"(\[(\d+),\d+\])", group = 1)),
          k = as.numeric(str_extract(variable, r"(\[\d+,(\d+)\])", group = 1))
        )
    }

    if (str_detect(extract, "k")) {
      df <-
        df |>
        left_join(
          party,
          by = join_by(k == party_stan),
          relationship = "many-to-one"
        )
    }
  }

  return(df)
}

pri_post_summary <- function(prior_, posterior_, party_map_, extract = "tk") {
  prior <- NULL
  if (!is.null(prior_)) {
    prior <- summarise_samples(
      x = prior_,
      party = party_map_,
      extract = extract
    )
    prior$type <- "prior"
  }

  posterior <- NULL
  if (!is.null(posterior_)) {
    posterior <- summarise_samples(
      x = posterior_,
      party = party_map_,
      extract = extract
    )
    posterior$type <- "posterior"
  }

  bind_rows(prior, posterior)
}

mu_raw <- pri_post_summary(mu_raw_pri, mu_raw_post, party_map, extract = "k")
mu <- pri_post_summary(mu_pri, mu_post, party_map, extract = "k")
beta <- pri_post_summary(beta_pri, beta_post, party_map, extract = "k")
sigma <- pri_post_summary(sigma_pri, NULL, party_map, extract = "k")
alpha_raw <- pri_post_summary(alpha_raw_pri, alpha_raw_post, party_map)
alpha <- pri_post_summary(alpha_pri, alpha_post, party_map)
```

Get draws.

```{r}
draw_samples <- function(x, party, extract = "tk") {
  df <-
    x |>
    as_draws_df() |>
    as_tibble() |>
    pivot_longer(-c(.chain, .iteration, .draw), names_to = "variable")

  if (!is.null(extract)) {
    if (extract == "t") {
      df <-
        df |>
        mutate(
          t = as.numeric(str_extract(variable, r"(\[(\d+)])", group = 1))
        )
    } else if (extract == "k") {
      df <-
        df |>
        mutate(
          k = as.numeric(str_extract(variable, r"(\[(\d+)\])", group = 1))
        )
    } else if (extract == "tk") {
      df <-
        df |>
        mutate(
          t = as.numeric(str_extract(variable, r"(\[(\d+),\d+\])", group = 1)),
          k = as.numeric(str_extract(variable, r"(\[\d+,(\d+)\])", group = 1))
        )
    }

    if (str_detect(extract, "k")) {
      df <-
        df |>
        left_join(
          party,
          by = join_by(k == party_stan),
          relationship = "many-to-one"
        )
    }
  }

  return(df)
}

pri_post_draws <- function(prior_, posterior_, party_map_, extract = "tk") {
  prior <- NULL
  if (!is.null(prior_)) {
    prior <- draw_samples(prior_, party_map_, extract = extract)
  }
  prior$type <- "prior"

  posterior <- NULL
  if (!is.null(posterior_)) {
    posterior <- draw_samples(posterior_, party_map_, extract = extract)
  }
  posterior$type <- "posterior"

  bind_rows(prior, posterior)
}


mu_raw_draws <- pri_post_draws(
  mu_raw_pri,
  mu_raw_post,
  party_map,
  extract = "k"
)

mu_draws <- pri_post_draws(
  mu_pri,
  mu_post,
  party_map,
  extract = "k"
)

beta_draws <- pri_post_draws(
  beta_pri,
  beta_post,
  party_map,
  extract = "k"
)

sigma_draws <- pri_post_draws(
  sigma_pri,
  NULL,
  party_map,
  extract = "k"
)

alpha_raw_draws <- pri_post_draws(
  alpha_raw_pri,
  alpha_raw_post,
  party_map
)

alpha_draws <- pri_post_draws(
  alpha_pri,
  alpha_post,
  party_map
)
```

## Quality of chains

### Mu raw

```{r}
mu_raw |>
  select(type, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(aes(x = k, y = value, color = type, shape = type), size = 3) +
  facet_wrap(facets = vars(name), ncol = 1, scale = "free") +
  scale_x_parties(party_map) +
  scale_shape_type() +
  scale_color_type() +
  theme_bw() +
  labs(
    title = "Sampling statistics for parameter: mu_raw"
  )
```

### Mu

```{r}
mu |>
  select(type, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(aes(x = k, y = value, color = type, shape = type), size = 3) +
  facet_wrap(facets = vars(name), ncol = 1, scale = "free") +
  scale_x_parties(party_map) +
  scale_shape_type() +
  scale_color_type() +
  theme_bw() +
  labs(
    title = "Sampling statistics for parameter: mu"
  )
```

### Beta

```{r}
beta |>
  select(type, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(aes(x = k, y = value, color = type, shape = type), size = 3) +
  facet_wrap(facets = vars(name), ncol = 1, scale = "free") +
  scale_x_parties(party_map) +
  scale_shape_type() +
  scale_color_type() +
  theme_bw() +
  labs(
    title = "Sampling statistics for parameter: Beta"
  )
```

### Sigma

```{r}
sigma |>
  select(type, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(aes(x = k, y = value, color = type, shape = type), size = 3) +
  facet_wrap(facets = vars(name), ncol = 1, scale = "free") +
  scale_x_parties(party_map) +
  scale_shape_type() +
  scale_color_type() +
  theme_bw() +
  labs(
    title = "Sampling statistics for parameter: Sigma"
  )
```

### Alpha raw (log scale)

```{r}
alpha_raw |>
  select(type, t, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(
    mapping = aes(
      x = t,
      y = value,
      shape = type,
      color = party_shortcut,
      group = party_shortcut
    )
  ) +
  scale_color_parties(party_map) +
  geom_vline(
    xintercept = unique(data_polls$t),
    linetype = "dashed",
    color = "black",
    alpha = 0.3
  ) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  scale_x_polls(cal) +
  scale_shape_type() +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Statistics for alpha_raw",
    x = NULL,
    y = NULL
  )
```

### Alpha

```{r}
alpha |>
  select(type, t, k, party_shortcut, rhat, ess_bulk, ess_tail) |>
  filter(k != max(k)) |>
  pivot_longer(
    cols = c(rhat, ess_bulk, ess_tail)
  ) |>
  ggplot() +
  geom_point(
    mapping = aes(
      x = t,
      y = value,
      shape = type,
      color = party_shortcut,
      group = party_shortcut
    )
  ) +
  scale_color_parties(party_map) +
  geom_vline(
    xintercept = unique(data_polls$t),
    linetype = "dashed",
    color = "black",
    alpha = 0.3
  ) +
  facet_grid(rows = vars(name), cols = vars(type), scales = "free") +
  scale_x_polls(cal) +
  scale_shape_type() +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Statistics for alpha",
    x = NULL,
    y = NULL
  )
```

## Distributions

### Mu_raw

```{r}
mu_raw_draws |>
  filter(k != max(k, na.rm = TRUE)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: mu_raw",
    x = NULL,
    y = NULL
  )
```

### Mu

```{r}
mu_draws |>
  filter(k != max(k, na.rm = TRUE)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: mu",
    x = NULL,
    y = NULL
  )
```

### Beta

```{r}
beta_draws |>
  filter(!is.na(k)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: beta",
    x = NULL,
    y = NULL
  )
```

### Sigma

```{r}
sigma_draws |>
  filter(!is.na(k)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: sigma",
    x = NULL,
    y = NULL
  )
```

### Alpha raw (log scale)

```{r}
alpha_raw_draws |>
  filter(k != max(k)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: alpha_raw",
    x = NULL,
    y = NULL
  )
```

### Alpha

```{r}
alpha_draws |>
  filter(k != max(k)) |>
  ggplot() +
  geom_density(
    mapping = aes(
      x = value,
      fill = type
    ),
    alpha = 0.5
  ) +
  facet_wrap(facets = vars(party_shortcut), scales = "free") +
  scale_fill_type() +
  theme_bw() +
  labs(
    title = "Distributions for parameter: alpha",
    x = NULL,
    y = NULL
  )
```

## Posterior checks

### Mu

```{r}
ggplot() +
  geom_point(
    data = cal,
    mapping = aes(x = t, y = voteshare, color = party_shortcut)
  ) +
  geom_hline(
    data = filter(mu, type == "posterior", !is.null(k)),
    mapping = aes(yintercept = mean, color = party_shortcut)
  ) +
  geom_rect(
    data = filter(mu, type == "posterior", !is.null(k)),
    mapping = aes(
      xmin = -Inf,
      xmax = Inf,
      ymin = `5.5%`,
      ymax = `94.5%`,
      fill = party_shortcut
    ),
    alpha = 0.2
  ) +
  scale_x_polls(cal) +
  theme_bw()
```

### Alpha

```{r}
ggplot() +
  geom_point(
    data = cal,
    mapping = aes(x = t, y = voteshare, color = party_shortcut)
  ) +
  geom_line(
    data = filter(alpha, type == "posterior", k != max(k)),
    mapping = aes(x = t, y = mean, color = party_shortcut)
  ) +
  geom_errorbar(
    data = filter(alpha, type == "posterior", k != max(k)),
    mapping = aes(
      x = t,
      ymin = `5.5%`,
      ymax = `94.5%`,
      color = party_shortcut
    )
  ) +
  scale_x_polls(cal) +
  scale_y_voteshare() +
  scale_color_parties(party_map) +
  theme_bw()
```

```{r}

```