---
title: Models
---

# Preparation

Load libraries.

```{r}
library(rstan)
library(bayesplot)
```

Configure rstan to run in parallel.

```{r setup, include=FALSE}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

# Simple VAR

- Ignores missing data.
- Ignores structure from agencies.
- Ignores length of the poll.
- Ignores time between polls.

$$
\begin{aligned}
y_t &\sim Dir(\phi_t \cdot s_t)            \\ % Data generating process
\phi_t &= \text{softmax}(x_t)              \\ % Transform latent variable to vector of probabilities
x_t &= \alpha + \beta x_{t-1} + \sigma_t \\ % Drift in time for latent variable
\alpha &\sim N(0, 1)                       \\ % Simplified prior for alpha - the long term drift
\beta_{ii} &\sim N(1, 0.1)                 \\ % Reaction to past "self"
\beta_{ij} &\sim N(0, 0.1)                 \\ % Reaction to past "others"
\epsilon_t &\sim N(0, 0.1)                 \\ % Random noise
s_t &= \frac{N_t}{\kappa}                  \\ % How much weight to give to every poll
\kappa &\sim \text{Gamma}(4, 2)            \\ % Design effect - how imprecise is it
\end{aligned}
$$


## Predictive Prior Checking

Create artificial data in CLR form.

$\alpha$ - support of a given party in population
  - $\alpha > 0$ - increases in popularity
  - $\alpha < 0$ - decreases in popularity

$\beta_{ii}$ - how does past affect the partys current polling (*memory*)
  - $\beta_{ii} > 1$ - explosion, increase in party polling further increases it's support
  - $\beta_{ii} \approx 1$ - party is mainly driven by random shocks
  - $\beta_{ii} < 1$ - party experiences short term bounces, but goes back to it's support

$\sigma$ - news sensitivity, surprise

```{r}
T <- 80
x_start <- c(0.35, 0.30, 0.20, 0.15)
alpha_input <- c(0.001, 0.000, 0.000, -0.001)
P <- length(alpha_input)
beta_input <- diag(c(0.98, 0.98, 0.90, 0.85))
# beta_input[1, 2] <- -0.03
# beta_input[3, 2] <- 0.03
sigma_input <- c(0.1, 0.02, 0.05, 0.05)
kappa_input <- 2.0

set.seed(789)
make_clr_matrix <- function(mat) {
  P <- nrow(mat)
  new_mat <- mat
  row_means <- rowMeans(mat)
  col_means <- colMeans(mat)
  grand_mean <- mean(mat)

  for (i in 1:P) {
    for (j in 1:P) {
      new_mat[i, j] <- mat[i, j] - row_means[i] - col_means[j] + grand_mean
    }
  }
  return(new_mat)
}
stan_data <- list(
  T = T,
  P = P,
  N = round(rnorm(T, 1500, 50)),
  x_start = x_start,
  sim_alpha = alpha_input,
  sim_beta = make_clr_matrix(beta_input),
  sim_sigma = sigma_input,
  kappa = kappa_input
)
```

```{r}
model_ppc <- stan(
  model_name = "Simple VAR - PPC",
  file = "models/simple_var_ppc.stan",
  data = stan_data,
  iter = 4000,
  warmup = 2000,
  chains = 1,
  open_progress = FALSE,
  algorithm = "Fixed_param"
)

saveRDS(model_ppc, file = config::get("simple_var_ppc", config = "gold"))

print(model_ppc@model_pars)
model_ppc
```

### Trace plots

Intercept.

```{r}
model_ppc |>
  extract(inc_warmup = TRUE, permuted = FALSE) |>
  mcmc_trace(
    regex_pars = "alpha_clr",
    n_warmup = 2000
  )
```

Beta.

```{r}
model_ppc |>
  extract(inc_warmup = TRUE, permuted = FALSE) |>
  mcmc_trace(
    regex_pars = "beta",
    n_warmup = 2000,
    facet_args = list(
      ncol = 3,
      nrow = 3
    )
  )
```

Kappa.

```{r}
model_ppc |>
  extract(inc_warmup = TRUE, permuted = FALSE) |>
  mcmc_trace(
    regex_pars = "kappa",
    n_warmup = 2000
  )
```

Sigma.

```{r}
model_ppc |>
  extract(inc_warmup = TRUE, permuted = FALSE) |>
  mcmc_trace(
    regex_pars = "sigma",
    n_warmup = 2000
  )
```

Latent variable x.

```{r}
# TODO...
```

Phi.

```{r}
# TODO...
```

Beta (CLR).

```{r}
model_ppc |>
  extract(inc_warmup = TRUE, permuted = FALSE) |>
  mcmc_trace(
    regex_pars = "beta_clr",
    n_warmup = 2000
  )
```

### Posteriors

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(rstan)

# 1. Extract the Simulation Data
# output structure: [Iterations, Days, Parties]
y_sim <- rstan::extract(model_ppc, pars = "y_sim")$y_sim

# Get dimensions
n_days <- dim(y_sim)[2]
n_parties <- dim(y_sim)[3]

# 2. Process Data for Each Party
plot_data_list <- list()

for (p in 1:n_parties) {
  # Extract matrix for party p: [Iter, Days]
  mat_p <- y_sim[,, p]

  # Calculate summary statistics for every day
  df_p <- data.frame(
    Day = 1:n_days,
    Party = paste0("Party ", p),
    Median = apply(mat_p, 2, median),
    L50 = apply(mat_p, 2, quantile, probs = 0.25),
    U50 = apply(mat_p, 2, quantile, probs = 0.75),
    L90 = apply(mat_p, 2, quantile, probs = 0.05),
    U90 = apply(mat_p, 2, quantile, probs = 0.95)
  )
  plot_data_list[[p]] <- df_p
}

# Combine into one main dataframe
df_all <- do.call(rbind, plot_data_list)

# 3. Plot All Parties
ggplot(df_all, aes(x = Day, group = Party, fill = Party, color = Party)) +
  # 90% Uncertainty Interval (Lightest transparency)
  geom_ribbon(aes(ymin = L90, ymax = U90), alpha = 0.15, color = NA) +

  # 50% Uncertainty Interval (Medium transparency)
  geom_ribbon(aes(ymin = L50, ymax = U50), alpha = 0.3, color = NA) +

  # Median Trend Line (Solid)
  geom_line(aes(y = Median), size = 1) +

  # Styling
  theme_minimal() +
  labs(
    title = "Projected Vote Share (All Parties)",
    subtitle = "Median trajectory with 50% and 90% credible intervals",
    y = "Vote Share",
    x = "Days"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.6)) + # Adjust limit as needed
  scale_fill_brewer(palette = "Set1") + # Nice distinct colors
  scale_color_brewer(palette = "Set1")
```
